Aligning marine species range data to better serve science and conservation

### 2016-02-06 ms_2_intro_res_disc.Rmd
blarg

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of species.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` species).  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align, determine several issues that lead to misalignment between predicted species distributions, and outline possible improvements. <!--BSH: I think Jamie makes this comment below, but we may need to be able to substantiate this somehow.  Is there a way to quickly figure out how many studies have used AM vs. IUCN?-->

@@

The two datasets have notably different taxonomic (Fig. 1A) and regional (Fig. 1B,C) coverage. IUCN-mapped species focus on tropical latitudes and away from the Atlantic and Eastern Pacific compared to AquaMaps-mapped species. These differences can be mitigated by using both datasets, but the underlying methodological differences complicate such direct comparisons. <!--_and furthermore requires resolving differences (or choosing one source over the other) for the `r format(n_spp_both, big.mark = ',')` shared species._-->

@@

<!--BSH: I think it’s stronger to start with the negative relationship overall, then talk about the 4 quadrants.  So I rearranged, in turn requiring a bit of editing-->
The negative linear relationship <!--BSH: Any stats for this? If so, insert just after this phrase. Otherwise maybe phrase it as 'The general pattern that emerges of a negative linear relationship, where increasing similarity...'--> indicates that increasing similarity in extent correlates with decreasing distribution alignment. <!--BSH: Could also be said as increasing similarity in alignment leads to decreasing similarity in extent.  The phrasing implies causality, even though you are careful to use ‘correlates’--> AquaMaps tends to extrapolate species ranges beyond known occurrences, such that the marginal range predicted by AquaMaps will fall in different locations than the marginal range predicted using IUCN methodology. For species with dissimilar extents, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar extents, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

@@

The extent-mismatched ranges contained in the upper left quadrant (n = `r n_q2`) include many species whose spatial distribution is similar, but where the IUCN range is notably larger, often extending into deeper water.  For example, corals dominate this quadrant (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), and IUCN range maps tend to extend corals into waters beyond their depth tolerance. <!-- BSH: Worth a few spot checks to confirm this is more or less true. And then maybe include 2-3 examples as figures in the SOM and refer to them here.-->
Ocean depth is explicitly included in AquaMaps models, while depth is frequently overlooked as a factor in IUCN range maps. _BSH: Simply clipping IUCN range maps to known depth preferences would resolve many of these mismatches._ <!--BSH: This is kinda redundant, just emphasizing a solution to the problem.  Not sure it's needed, or perhaps could be woven in to previous sentence.--> _commission error terminology?_

Species found in the lower right quadrant (n = `r n_q3`) often represent cases of "two wrongs make a right."  For these species, IUCN ranges frequently overextend into unsuitable depths, as in the case of many upper left quadrant species, while at the same time AquaMaps ranges often aggressively extrapolate presence into locations where IUCN predicts absence.  Consequently, area ratios are close to 100%, but similar extents are meaningless when the distributions are poorly aligned. _commission error terminology?_

The most vexing cases are in the lower left quadrant (n = `r n_q4`), where neither distribution nor extent match.  Often these mismatches arise when AquaMaps relies on limited data to predict species presence well beyond known occurrences.  Indeed, `r p_q4_datapoor`% (n = `r n_q4_datapoor`) of these cases are data-poor (in which the AquaMaps environmental envelope model is based on fewer than 10 known cell occurrences), compared with `r p_q123_datapoor`% for species in the other three quadrants.  Only `r p_q4_rev`% (n = `r n_q4_rev`) of AquaMaps maps in this quadrant have undergone expert review, compared with `r p_q123_rev`% of species in the other three quadrants. <!--BSH: This implies the problem is solely with AM.  Are there no issues with IUCN maps that cause problems in this quadrant?--> ___The data-poor condition likely applies to IUCN as well - AM using OBIS points to generate models, and IUCN is probably doing the same.  Check, and rework.  This means AM is creating errors of commission, while IUCN may be committing errors of omission for poorly-studied species and/or regions - compounding the total problem.___

<!-- OLD: This breakdown by quadrant helps explain the negative linear relationship seen in the plot, in which increasing similarity in extent correlates with decreasing distribution alignment.  When extrapolating species range beyond known occurrences, the marginal range predicted by the AquaMaps model will fall in different locations than the marginal range predicted using IUCN methodology.  For species with dissimilar extents, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range.  As we examine species with increasingly similar extents, the differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.-->

>>>>

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of species.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` species).  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align, determine several issues that lead to misalignment between predicted species distributions, and outline possible improvements. <!--BSH: I think Jamie makes this comment below, but we may need to be able to substantiate this somehow.  Is there a way to quickly figure out how many studies have used AM vs. IUCN?  CCO: What about age - AquaMaps is pretty recent; before that, would IUCN have been the dominant dataset for marine?  Maybe we can track citations of AquaMaps, but citations of IUCN I think are generic - just cite the Red List - not referencing spatial information specifically, much less marine.-->

@@

The two datasets have notably different taxonomic (Fig. 1A) and regional (Fig. 1B,C) coverage. IUCN-mapped species focus on tropical latitudes and away from the Atlantic and Eastern Pacific compared to AquaMaps-mapped species. These differences can be mitigated by using both datasets, but the underlying methodological differences complicate such direct comparisons.

@@

The negative linear relationship <!--BSH: Any stats for this? If so, insert just after this phrase. Otherwise maybe phrase it as 'The general pattern that emerges of a negative linear relationship, where increasing similarity...'--> indicates that increasing similarity in extent correlates with decreasing distribution alignment. <!--BSH: Could also be said as increasing similarity in alignment leads to decreasing similarity in extent.  The phrasing implies causality, even though you are careful to use ‘correlates’--> AquaMaps tends to extrapolate species ranges beyond known occurrences, such that the marginal range predicted by AquaMaps will fall in different locations than the marginal range predicted using IUCN methodology. For species with dissimilar extents, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar extents, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

@@

The extent-mismatched ranges contained in the upper left quadrant (n = `r n_q2`) include many species whose spatial distribution is similar, but where the IUCN range is notably larger, often extending into deeper water.  For example, corals dominate this quadrant (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), and IUCN range maps tend to extend corals into waters beyond their preferred depths, likely introducing errors of commission. <!-- BSH: Worth a few spot checks to confirm this is more or less true. And then maybe include 2-3 examples as figures in the SOM and refer to them here.-->
Ocean depth is explicitly included in AquaMaps models, while depth is frequently overlooked as a factor in IUCN range maps. _BSH: Simply clipping IUCN range maps to known depth preferences would resolve many of these mismatches._ <!--BSH: This is kinda redundant, just emphasizing a solution to the problem.  Not sure it's needed, or perhaps could be woven in to previous sentence.-->

Species found in the lower right quadrant (n = `r n_q3`) often represent cases of "two wrongs make a right."  For these species, IUCN ranges frequently overextend into unsuitable depths, as in the case of many upper left quadrant species, while at the same time AquaMaps ranges often aggressively extrapolate presence into locations where IUCN predicts absence, introducing additional commission errors.  Consequently, area ratios are close to 100%, though similar extents are meaningless when the distributions are poorly aligned.

<!--The most vexing cases are in the lower left quadrant (n = `r n_q4`), where neither distribution nor extent match.  Often these mismatches arise when AquaMaps relies on limited data to predict species presence well beyond known occurrences.  Indeed, `r p_q4_datapoor`% (n = `r n_q4_datapoor`) of these cases are data-poor (in which the AquaMaps environmental envelope model is based on fewer than 10 known cell occurrences), compared with `r p_q123_datapoor`% for species in the other three quadrants.  Only `r p_q4_rev`% (n = `r n_q4_rev`) of AquaMaps maps in this quadrant have undergone expert review, compared with `r p_q123_rev`% of species in the other three quadrants. <!--BSH: This implies the problem is solely with AM.  Are there no issues with IUCN maps that cause problems in this quadrant?--> ___The data-poor condition likely applies to IUCN as well - AM using OBIS points to generate models, and IUCN is probably doing the same.  Check, and rework.  This means AM is creating errors of commission, while IUCN may be committing errors of omission for poorly-studied species and/or regions - compounding the total problem.___-->

The most vexing cases are in the lower left quadrant (n = `r n_q4`), where neither distribution nor extent match.  When extrapolating from limited observations, the AquaMaps model often predicts species presence well beyond known occurrences; at the same time, IUCN range maps .  Indeed, `r p_q4_datapoor`% (n = `r n_q4_datapoor`) of these cases are data-poor (in which the AquaMaps environmental envelope model is based on fewer than 10 known cell occurrences ___is there a good way to get the number of OBIS observations for each of these species, as an external validation of data-poor?)___, compared with `r p_q123_datapoor`% for species in the other three quadrants.  Only `r p_q4_rev`% (n = `r n_q4_rev`) of AquaMaps maps in this quadrant have undergone expert review, compared with `r p_q123_rev`% of species in the other three quadrants. <!--BSH: This implies the problem is solely with AM.  Are there no issues with IUCN maps that cause problems in this quadrant?--> ___The data-poor condition likely applies to IUCN as well - AM using OBIS points to generate models, and IUCN is probably doing the same.  Check, and rework.  This means AM is creating errors of commission, while IUCN may be committing errors of omission for poorly-studied species and/or regions - compounding the total problem.___
